---
output:
  html_document:
    fig_caption: yes
---
# Reproducible Research: Peer Assessment 1
Author: girishbabu
Version: 08/16/2014

Load the ggplot2 & data.table libraries to generate the plots

```{r, echo=TRUE}
library(ggplot2)
library(data.table)
```

## Loading and preprocessing the data

Extract the activity.zip file with the date and load to a data.frame object. Extract the step counts as numeric and intervals as date  

```{r, echo=TRUE}
df.activity = read.csv(unz("activity.zip", "activity.csv"), header=T,
                sep=",", stringsAsFactors=F, na.strings="NA",
                colClasses=c("numeric", "Date", "numeric"))

dt.activity = data.table(df.activity)
rm(df.activity)
dt.activity
```

## What is mean total number of steps taken per day?

Combine the steps per day

```{r, echo=TRUE}
dt.totalsteps = dt.activity[, sum(steps, na.rm=T), by="date"]
head(dt.totalsteps)
```

Create a histogram with the total steps data, and calculate their mean and median values

```{r, echo=TRUE}
qplot(dt.totalsteps$V1, geom="histogram", binwidth=1000,
      main="Total number of steps per day", xlab="Steps", fill=..count..)
median(dt.totalsteps$V1)
mean(dt.totalsteps$V1)
```

## What is the average daily activity pattern?

Combine the steps by each interval that we have available in the dataset

```{r, echo=TRUE}
dt.pattern = dt.activity[, sum(steps, na.rm=T)/.N, by="interval"]
head(dt.pattern)
```

Plot the pattern of daily activity averaged over all days

```{r, echo=TRUE}
qplot(interval, V1, data = dt.pattern, xlab="Interval", ylab="Number of Steps",
       color=I("red"),  geom = "line", main="Daily Activity Pattern")
```

We can also figure out which interval is the one that contains the maximum number
of stepts across all the days.

```{r, echo=TRUE}
dt.pattern[V1 == max(dt.pattern$V1)]$interval
```

## Imputing missing values

First, let us verify what is the number of missing values in the dataset.

```{r, echo=TRUE}
sum(is.na(dt.activity$steps))
```

We will fill these missing vaules by using the mean of the specific 5-minute interval
as a replacement for the missing value. I already have these quantities calculated
in the `dt.pattern` table.

```{r, echo=TRUE}
dt.nomissing = copy(dt.activity)
dt.nomissing[is.na(steps), steps := dt.pattern[dt.pattern$interval == interval]$V1]
sum(is.na(dt.nomissing$steps))
```

Now, we again create an histogram with the total steps data, and calculate their *mean* 
and *median* values to verify the changes we made on the dataset.

```{r, echo=TRUE}
dt.totalsteps.noNA = dt.nomissing[, sum(steps, na.rm=T), by="date"]
head(dt.totalsteps.noNA)
qplot(dt.totalsteps.noNA$V1, geom="histogram", binwidth=3000,
      main="Total Number of Steps per Day (no NAs)", xlab="Steps", fill=..count..)
median(dt.totalsteps.noNA$V1)
mean(dt.totalsteps.noNA$V1)
```

As a result from this change, the mean got really close to the median, and the
shape of the histogram got much closer to a bell shape. The concentration of
days with less then 3000 steps got really small, as most of them were filled with
missing values, and that went to the bucket that contains the mean and median.

## Are there differences in activity patterns between weekdays and weekends?

First, let us separate our dates in weekdays and weekends so we can start splitting
the interval counts among them. For this we will create a factor variable and
add it to the dataset

```{r, echo=TRUE}
activity.weekdays = weekdays(dt.activity$date)
factor.weekday = factor( ifelse(activity.weekdays == "Saturday" 
                                | activity.weekdays == "Sunday",
                                "weekend",
                                "weekday"),
                         levels = c("weekend","weekday"))
dt.activity[, weekstatus := factor.weekday]
```

We then proceed to do some aggregating magic with our data.table

```{r, echo=TRUE}
dt.pattern.week = dt.activity[, sum(steps, na.rm=T)/.N, by=c("weekstatus", "interval")]
head(dt.pattern.week)
```

Now, we create a plot to help us visualize the pattern of daily activity averaged
over weekends and weekdays.

```{r, echo=TRUE}
qplot(interval, V1, data = dt.pattern.week, facets = weekstatus ~ .,
      xlab="Interval", ylab="Number of Steps",
      color=I("red"),  geom = "line", main="Daily Activity Pattern")
``````